Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-08-14 Paper/app.tex
..

Several useful inferences can be drawn from the above plot. Firstly, the total Sobol indices are found to be
consistent
with the normalized activity scores evaluated uing the gradient-based approach. Whereas, the gradient-free
approach yields consistent results for the $A_i$'s, it does not capture the sensitivity associated with
$E_{a,15}$. While this observation bolsters our confidence in the gradient-based approach, it clearly
underscores the shortcoming of the gradient-free approach. Numerical errors are incurred when approximating
the model gradients using the local linear approximation in this case. In~\ref{sub:verify}, we assess the
suitability of the gradient-free approach in quantifying the uncertainty in the ignition delay. 
Secondly, our findings indicate that the  
the ignition delay is predominantly sensitivity towards $A_1$, $A_9$, $E_{a,1}$, and $E_{a,15}$ and 
moderately sensitivity towards $A_{15}$ and $A_{17}$. Sensitivity towards the remaining uncertain inputs is
found to be low or negligible. Note that his observation can also be exploited for reducing the 
dimensionality of the problem from 33 inputs to 4-6 inputs depending upon the required level of accuracy.
The sensitivity-driven dimension reduction for uncertainty quantification has been
discussed in our earlier effort~\cite{Vohra:2018}. In this work, however, we focus on dimension reduction using the active 
subspace approach which is shown to yield even greater scope for dimension reduction i.e. from 33 to 1. Furthermore, we have
shown that the dominant eigenspace can be used to approximate global sensitivity measures with no additional effort.
The accuracy of the 1-dimensional surrogate, $\tilde{G}$, obtained using the two approaches is assessed in the following section.

@@

The 1-dimensional surrogate ($\tilde{G}$) is investigated for its accuracy as well as the ability to capture the 
uncertainty in the
model output in two ways. Firstly, we estimate the relative L-2 norm of the discrepancy~($\varepsilon_d$)
between estimates of ignition delay in the case of H$_2$/O$_2$ reaction, obtained using the 
model output and the surrogate in the following equation:
%
\be
\varepsilon_d = \frac{\|G(\bm{\xi}) - \tilde{G}(\bm{\xi})\|_2}{\|G(\bm{\xi})\|_2}
\ee
%
The relative norm, $\varepsilon_d$ was estimated to be 8.50$\times10^{-2}$ based on a cross-validation test data
comprising 10$^{4}$ model evaluations. Hence, in the norm-sense, it can be said that the 1-dimensional surrogate is 
reasonably accurate.

Secondly, we verify the accuracy of the surrogate in a probabilistic setting. In particular, we compare a histogram
plot of the model evaluations (in the cross-validation set) with a PDF evaluated on the same set of samples
using the surrogate, $\tilde{G}$ in Figure~\ref{fig:pdf_33D}. The PDF is evaluated using surrogates constructed for
a 1-dimensional as well as a 2-dimensional active subspace. 
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.45\textwidth]{./Figures/pdf_comp_id_1e4}
\caption{A comparison of the histogram plot based on model evaluations and the PDF based on surrogate
predictions for ignition delay. The two plots are based on samples in the validation set.}
\label{fig:pdf_33D}
\end{center}
\end{figure}
%
In the above comparison, it is observed that the two PDFs are almost identical, thereby confirming a 1-dimensional
active subspace. The modal estimate from either PDF is observed to be in agreement with that
based on the histogram plot. The mean value based on the set of model evaluations used for verification as well
as the 1-dimensional surrogate was found to be 0.133, whereas, the standard deviations were found to be 0.0198 and 0.0160
respectively. Hence, while the surrogate yields an accurate estimation of the mean, the standard deviation is
under-estimated. This can be understood from the SSP in Figure~\ref{fig:hd} (right) which shows that the model
evaluations do not exactly lie on the linear 1-dimensional surrogate used to capture them. Therefore, the
1-dimensional surrogate can be used to estimate the mean as well as the modal estimate of the ignition delay in
the presence of uncertainty in the 33 parameter considered in this application. The uncertainty associated with the
ignition delay though reasonably close to the true value, is under-estimated.

>>>>

Several useful inferences can be drawn from the above plot. Firstly, the total Sobol indices are found to be
consistent
with the normalized activity scores evaluated uing the gradient-based approach. Whereas, the gradient-free
approach yields consistent results for the $A_i$'s, it does not capture the sensitivity associated with
$E_{a,15}$. While this observation bolsters our confidence in the gradient-based approach, it clearly
underscores the shortcoming of the gradient-free approach. Numerical errors are incurred when approximating
the model gradients using the local linear approximation in this case. In~\ref{sub:verify}, we further assess the
suitability of the gradient-free approach in quantifying the uncertainty in the ignition delay. 
Secondly, our findings indicate that   
the ignition delay is predominantly sensitive towards $A_1$, $A_9$, $E_{a,1}$, and $E_{a,15}$ and 
moderately sensitive towards $A_{15}$ and $A_{17}$. Sensitivity towards the remaining uncertain inputs is
found to be low or negligible. Note that this observation can also be exploited for reducing the 
dimensionality of the problem from 33 inputs to 4-6 inputs depending upon the required level of accuracy.
The sensitivity-driven dimension reduction for uncertainty quantification has been
discussed in our earlier effort~\cite{Vohra:2018}. In this work, however, we focus on dimension reduction using the active 
subspace approach which is shown to yield even greater scope for dimension reduction i.e. from 33 to 1. Furthermore, we have
shown that the dominant eigenspace can be used to approximate global sensitivity measures with no additional effort.
The accuracy of the 1-dimensional surrogate, $\tilde{G}$, obtained using the two approaches is assessed in the following section.

@@

The 1-dimensional surrogate ($\tilde{G}$) is investigated for its accuracy as well as the ability to capture the 
uncertainty in the
model output in two ways. Firstly, we estimate the relative L-2 norm of the discrepancy~($\varepsilon_d$)
between estimates of ignition delay in the case of H$_2$/O$_2$ reaction, obtained using the 
model output and the surrogate in the following equation:
%
\be
\varepsilon_d = \frac{\|G(\bm{\xi}) - \tilde{G}(\bm{\xi})\|_2}{\|G(\bm{\xi})\|_2}
\ee
%
The relative norm, $\varepsilon_d$ was estimated to be 1.35$\times10^{-2}$ and 1.02$\times10^{-1}$
using $\tilde{G}$ from the gradient-based approach and the gradient-free approach respectively. Hence,
an order of magnitude increase in $\varepsilon_d$ was estimated in the case of gradient-free approach.
Model evaluations ($G$) at 10$^{4}$ samples in the input domain were used for this purpose. 
Hence, in the norm-sense, it can be said that the 1-dimensional surrogate in the active subspace, obtained
using the gradient-based approach is remarkably accurate with a relative error of less than 1.5$\%$.

Secondly, we verify the accuracy of the two surrogates in a probabilistic setting. In particular, we compare 
probability density functions (PDFs) obtained using the true set of model evaluations, and 1-dimensional
surrogates from the two approaches, in Figure~\ref{fig:pdf_33D}. Note that the three PDFs were evaluated 
using the same set of 10$^4$ samples in the cross-validation set. 
%
\begin{figure}[htbp]
\begin{center}
\begin{minipage}[htbp]{.25\linewidth}
\vspace{0pt}
%\centering
\hspace{-25mm}
\begin{tabular}{ccc}
\toprule
$\textbf{Distribution}$ & $\mu$ & $\sigma$ \\ 
\bottomrule
$G$~(Model) & 0.133 & 0.0198 \\
$\tilde{G}$~(Gradient-based) & 0.133 & 0.0196 \\
$\tilde{G}$~(Gradient-free) & 0.133 & 0.0167 \\
\bottomrule
\end{tabular}
\end{minipage}
\hspace{5mm}
\begin{minipage}[htbp]{.25\linewidth}
\vspace{0pt}
%\centering
\includegraphics[width=3.0in]{./Figures/pdf_comp_id_1e4}
\end{minipage}%
\end{center} 
\caption{Left: Table providing the mean ($\mu$) and the standard deviation ($\sigma$) corresponding to
the three PDFs using evaluations
at 10$^4$ samples in the cross-validation set. Right: A comparison of the PDFs of ignition delay, obtained using model 
evaluations (solid line) and 1-dimensional surrogates using the gradient-free approach (dashed line) and the gradient-based
approach (dashed-dotted line). The same set of 10$^4$ samples in the cross-validation set were used in each case.}
\label{fig:pdf_33D}
\end{figure}
%
%
%\begin{figure}[htbp]
% \begin{center}
%  \includegraphics[width=0.45\textwidth]{./Figures/pdf_comp_id_1e4}
%\caption{A comparison of the PDFs of ignition delay, obtained using model evaluations (solid line) and
%1-dimensional surrogates using the gradient-free approach (dashed line) and the gradient-based approach
%(dashed-dotted line). The same set of 10$^4$ samples in the cross-validation set were used in each case.}
%\label{fig:pdf_33D}
%\end{center}
%\end{figure}
%
The PDFs based on the model evaluations and the gradient-based approach are nearly identical, confirming
the existence of a 1-dimensional active subspace as well as the accuracy of the gradient-based approach. 
Whereas, the PDF based on the gradient-free approach is observed to show consistency with the true PDF in the modal
estimate; it underestimates the spread in the ignition delay. Specific values of the mean and the standard deviation
in each case are provided in the table in Figure~\ref{fig:pdf_33D}. Hence, the three PDFs are found to be
consistent in their prediction of mean value of the ignition delay.  The standard deviation estimates based on model
evaluations and the gradient-based approach are found to be in close agreement, whereas, it  is accurately estimated
by the gradient-free approach only upto the first significant figure.

### 2018-08-14 Paper/method.tex
..

The gradient-free approach when implemented to the 19-dimensional H$_2$/O$_2$ reaction kinetics problem
with uncertain pre-exponents ($A_i$'s) also yielded a 1-dimensional active subspace. In Figure~\ref{fig:comp} (left),
we provide a comparison of the components of the dominant eigenvector, obtained using the two approaches.
Additionally, the quantity, $\delta \hat{\mat{W}}_{1,j}^{(r)}$ is also plotted in
Figure~\ref{fig:comp} (right) to illustrate convergence characteristics for the gradient-free approach.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.8\textwidth]{./Figures/eigv6}
\caption{Left: An illustrative comparison of individual components of the converged dominant eigenvector obtained
using the two approaches discussed in~\ref{sub:grad} and~\ref{sub:gradfree}. Right: The quantity,  $\delta \hat{\mat{W}}_{1,j}^{(r)}$
is plotted for successive iterations to illustrate the convergence behavior.}
\label{fig:comp}
\end{center}
\end{figure}
%
Individual components of the dominant eigenvector from the two approaches are observed to be in excellent
agreement with each other. Moreover, as expected,  $\delta \hat{\mat{W}}_{1,j}^{(r)}$ is observed to decrease with iterations, and
the eigenvector is considered to have converged if its value is found to be smaller than 0.1. Note that only 30 samples
were found to be sufficient for approximating the active subspace in the 19-dimensional input domain.

As mentioned earlier, the model output $f(\bm{\xi})$ predominantly varies in the active subspace. Hence, 
$f(\bm{\xi})$ can be approximated as $G(\hat{\mat{W}}_1^\top\bm{\xi})$ in
the subspace. The plot of $G$ versus $\hat{\mat{W}}_1^\top\bm{\xi}$, regarded as the \textit{sufficient summary plot} (SSP) is 
compared for the two approaches in Figure~\ref{fig:comp_ssp}.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.48\textwidth]{./Figures/comp_ssp}
\caption{An illustrative comparison of the SSPs generated using the gradient-based and the gradient-free approaches.}
\label{fig:comp_ssp}
\end{center}
\end{figure}
%
The two SSPs are observed to be in excellent agreement with each other. Moreover, it is interesting to note
that the response in ignition
delay based on the considered prior marginals for $A_i$'s is observed to be approximately linear in the
active subspace. This observation further indicates that the active subspace is 1-dimensional in this case.

@@

The above comparisons establish consistency between gradient-based and
gradient-free approaches. This indicates that the local linear approximation
of the model output resulted in a reasonably accurate estimation of the
gradients in this case. In the following section, we will once again use both
approaches to discover an active subspace in the high-dimensional 
problem involving 33 uncertain inputs.

>>>>

The gradient-free approach when implemented to the 19-dimensional H$_2$/O$_2$ reaction kinetics problem
with uncertain pre-exponents ($A_i$'s) also yielded a 1-dimensional active subspace. In Figure~\ref{fig:comp} (left),
we provide a comparison of the square of individual components of the dominant eigenvector, obtained using the two 
approaches.
Additionally, the quantity, $\delta \hat{\mat{W}}_{1,j}^{(r)}$ is also plotted in
Figure~\ref{fig:comp} (right) to illustrate convergence characteristics for the gradient-free approach.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.8\textwidth]{./Figures/eigv6}
\caption{Left: An illustrative comparison of individual squared components of the converged dominant eigenvector obtained
using the two approaches discussed in~\ref{sub:grad} and~\ref{sub:gradfree}. Right: The quantity,  $\delta \hat{\mat{W}}_{1,j}^{(r)}$
is plotted for successive iterations to illustrate the convergence behavior.}
\label{fig:comp}
\end{center}
\end{figure}
%
Individual components of the dominant eigenvector from the two approaches are observed to be in excellent
agreement with each other. Moreover, as expected,  $\delta \hat{\mat{W}}_{1,j}^{(r)}$ is observed to decrease with iterations, and
the eigenvector is considered to have converged if its value is found to be smaller than 0.1. Note that only 30 samples
were found to be sufficient for approximating the active subspace in the 19-dimensional input domain.

As mentioned earlier, the model output $f(\bm{\xi})$ predominantly varies in the active subspace. Hence, 
$f(\bm{\xi})$ can be approximated as $G(\hat{\mat{W}}_1^\top\bm{\xi})$ in
the subspace. The plot of $G$ versus $\hat{\mat{W}}_1^\top\bm{\xi}$, regarded as the \textit{sufficient summary plot} (SSP) is 
compared for the two approaches in Figure~\ref{fig:comp_ssp}.
%
\begin{figure}[htbp]
 \begin{center}
  \includegraphics[width=0.45\textwidth]{./Figures/comp_ssp}
\caption{An illustrative comparison of the SSPs generated using the gradient-based and the gradient-free approaches.}
\label{fig:comp_ssp}
\end{center}
\end{figure}
%
The two SSPs are observed to be in excellent agreement with each other. Moreover, it is interesting to note
that the response in ignition
delay based on the considered prior marginals for $A_i$'s is observed to be approximately linear in the
active subspace. This observation further indicates that the active subspace is 1-dimensional in this case.

@@

The above comparisons establish consistency between gradient-based and
gradient-free approaches. This indicates that the local linear approximation
of the model output resulted in a reasonably accurate estimation of the
gradients in this case. In the following section, we shift our focus to 
the high-dimensional H$_2$/O$_2$ reaction kinetics application involving
33 uncertain inputs.
